Supermartingales and Termination Analysis of Statistical PCF

### 2021-01-18 LICS21/SPCF.tex
Minor fixes

\subsection{Trace semantics}
\label{sec:sampling semantics}
This version of the reduction relation allows $\tsample$ to reduce to any number in $[0,1]$. 
To more precisely specify the probabilities, an additional argument is needed to determine the outcome of random samples. Let $ I := [0,1] \subset \mathbb{R} $, and define $\entrosp := I^{\mathbb{N}}$, with the Borel $\sigma$-algebra $\Sigma_\entrosp$ and the probability measure, denoted $\mu_{\entrosp}$, given by the limit of $1 \gets I \gets I^2 \gets \cdots$, where the maps are the projections that ignore the last element. Equivalently, a basis of measurable sets is $\prod_{i=0}^\infty X_i$ where $X_i$ are all Borel and all but finitely many are $I$, and $\mu_{\entrosp} (\prod_{i=0}^\infty X_i) := \prod_{i=0}^\infty \Leb(X_i)$, \changed[lo]{writing $\Leb$ for the Lebesgue measure.}
\changed[lo]{The maps $\pi_h:\entrosp \to I$ (projecting to the first element) and $\pi_t:\entrosp \to \entrosp$ (popping the first element) are then measurable.}
Following \cite{DBLP:conf/esop/CulpepperC17}, we call the probability space $(\entrosp, \Sigma_\entrosp, \mu_{\entrosp})$ the \emph{entropy space}, and elements of $\entrosp$ \emph{traces}.

@@

\iffalse
\lo{Alternatively, define the \emph{runtime of $M$} to be the random variable 
\[
T_M(s) := 
\begin{cases}
\min \set{n \mid \pi_0(\red^n(M, s)) \textrm{ is a value}} & \hbox{if $\red^\infty(M,s)$ is defined}\\
\infty & \hbox{otherwise}
\end{cases}
\]
Equivalently, we say that $M$ is \emph{almost surely terminating} (AST) if $T_M < \infty$ a.s.; 
and $M$ is \emph{positively almost surely terminating} (PAST) if $\expect{T_M} < \infty$.}
\fi
For example, the term 
\[
(\tY \lambda f, n: \tif{\tsample - 0.5 < 0}{n}{f (n+1)} ) \, \underline{0},
\] 
which generates a geometric distribution, terminates on the set $\entrosp \setminus [0.5,1]^\mathbb N$, which has measure 1, therefore it terminates almost surely, whereas 
\(
\tif{\tsample - 0.5 < 0}{\underline 0}{(\tY \lambda x. x) \, \underline 0}
\), 
which terminates on the set $\pi_h^{-1}[[0,0.5]]$, has probability 0.5 of failing to terminate.

>>>>

\subsection{Trace semantics}
\label{sec:sampling semantics}
This version of the reduction relation allows $\tsample$ to reduce to any number in $[0,1]$. 
To more precisely specify the probabilities, an additional argument is needed to determine the outcome of random samples. Let $ I := [0,1] \subset \mathbb{R} $, and define $\entrosp := I^{\mathbb{N}}$, with the Borel $\sigma$-algebra $\Sigma_\entrosp$. Equivalently, a basis of measurable sets is $\prod_{i=0}^\infty X_i$ where $X_i$ are all Borel and all but finitely many are $I$, and the probability measure $\mu_{\entrosp}$ is given by $\mu_{\entrosp} (\prod_{i=0}^\infty X_i) := \prod_{i=0}^\infty \Leb(X_i)$, \changed[lo]{writing $\Leb$ for the Lebesgue measure.}
\changed[lo]{The maps $\pi_h:\entrosp \to I$ (projecting to the first element) and $\pi_t:\entrosp \to \entrosp$ (popping the first element) are then measurable.}
Following \cite{DBLP:conf/esop/CulpepperC17}, we call the probability space $(\entrosp, \Sigma_\entrosp, \mu_{\entrosp})$ the \emph{entropy space}, and elements of $\entrosp$ \emph{traces}.

@@

\iffalse
\lo{Alternatively, define the \emph{runtime of $M$} to be the random variable 
\[
T_M(s) := 
\begin{cases}
\min \set{n \mid \pi_0(\red^n(M, s)) \textrm{ is a value}} & \hbox{if $\red^\infty(M,s)$ is defined}\\
\infty & \hbox{otherwise}
\end{cases}
\]
Equivalently, we say that $M$ is \emph{almost surely terminating} (AST) if $T_M < \infty$ a.s.; 
and $M$ is \emph{positively almost surely terminating} (PAST) if $\expect{T_M} < \infty$.}
\fi
For example, the term 
\[
(\tY \lambda f n: \tif{\tsample - 0.5 < 0}{n}{f (n+1)} ) \, \underline{0},
\] 
which generates a geometric distribution, terminates on the set $\entrosp \setminus [0.5,1]^\mathbb N$, which has measure 1, therefore it terminates almost surely, whereas 
\(
\tif{\tsample - 0.5 < 0}{\underline 0}{(\tY \lambda x. x) \, \underline 0}
\), 
which terminates on the set $\pi_h^{-1}[[0,0.5]]$, has probability 0.5 of failing to terminate.

### 2021-01-18 LICS21/antitone-proofs.tex
Broken reference fixed, and other minors.
Minor fixes

@@

\Cref{ex:non-affine recursion}
Let 
\[
\Xi := \tY \lambda f x. x \oplus_{2/3} f (f (x + 1))
\] 
A suitable sparse ranking function for $\Xi \, \underline 1$ would be $\Xi^n \, \underline i \mapsto 3 \, n$ (for $n \geq 0$), because $\Xi^n \, \underline i$ reduces to either $\Xi^{n+1} \, \underline{i+1}$ or $\Xi^{n-1} \, \underline{i+1}$, with probabilities $1/3$ and $2/3$. The value of $i$ does not actually matter for the progress of this recursion. It is basically another variant of the biased random walk, except that the relevant variable is the number of copies of $\Xi$, instead of a real number in the term.

@@

Again, this is not actually the correct reduction order, in that $F$ should be applied to its argument $n_i$ and reduced to a value before its second argument is expanded, but this would complicate the presentation, and this version can be justified by \Cref{thm:confluent ranking} instead. The reduction strategy implied here should be clear enough (assuming that, where it isn't specified, it just matches $\cbv$), but to be more precise, let $r(F \underline{n_0} (\dots(F\ \underline {n_{k-1}}\ X)\dots)) = @_2^k;\cbv(X)$, where $X$ is not a value or of the form $F\ \underline{n_k}\ Z$ for some non-value $Z$. All terms match this pattern for exactly one value of $k \geq 0$, except for values. Within $X$, there is necessarily a redex at $\cbv(X)$. The only conditions that $@_2^k;\cbv(X)$ is also a position of a redex in the whole term are that if the redex is a $\tsample$, it is not inside a $\lambda$ or $\tY$ on the right of an application, which is true because $\cbv$ never selects a redex inside of a $\lambda$ or $\tY$. After $n$ reaches $0$, the term is $F \underline{n_0} (\dots(F\ \underline {n_k}\ \underline x)\dots)$, then the innermost $F$ is evaluated completely in $\cbv$ order, therefore (by assumption) it terminates. Because there are no $\tY$-reductions in the evaluation of $F$, it can't reach any other term of the form $F\ \underline{n_k'}\ X$, therefore $k$ never changes until this whole subexpression reaches a number, at which point $k$ decreases by 1 and the next $F$ is evaluated.

>>>>

%\medskip

\subsubsection*{Non-affine recursion}

@@

\Cref{ex:non-affine recursion easy}
Let 
\[
\Xi := \tY \lambda f x. x \oplus_{2/3} f (f (x + 1))
\] 
A suitable sparse ranking function for $\Xi \, \underline 1$ would be $\Xi^n \, \underline i \mapsto 3 \, n$ (for $n \geq 0$), because $\Xi^n \, \underline i$ reduces to either $\Xi^{n+1} \, \underline{i+1}$ or $\Xi^{n-1} \, \underline{i+1}$, with probabilities $1/3$ and $2/3$. The value of $i$ does not actually matter for the progress of this recursion. It is basically another variant of the biased random walk, except that the relevant variable is the number of copies of $\Xi$, instead of a real number in the term.

\subsubsection*{Higher-order recursion}

@@

Again, this is not actually the correct reduction order, in that $F$ should be applied to its argument $n_i$ and reduced to a value before its second argument is expanded, but this would complicate the presentation, and this version can be justified by \Cref{thm:confluent ranking} instead. The reduction strategy implied here should be clear enough (assuming that, where it isn't specified, it just matches $\cbv$), but to be more precise, let $r(F \underline{n_0} (\dots(F\ \underline {n_{k-1}}\ X)\dots)) = @_2^k;\cbv(X)$, where $X$ is not a value or of the form $F\ \underline{n_k}\ Z$ for some non-value $Z$. All terms match this pattern for exactly one value of $k \geq 0$, except for values. Within $X$, there is necessarily a redex at $\cbv(X)$. The only conditions that $@_2^k;\cbv(X)$ is also a position of a redex in the whole term are that if the redex is a $\tsample$, it is not inside a $\lambda$ or $\tY$ on the right of an application, which is true because $\cbv$ never selects a redex inside of a $\lambda$. After $n$ reaches $0$, the term is $F \underline{n_0} (\dots(F\ \underline {n_k}\ \underline x)\dots)$, then the innermost $F$ is evaluated completely in $\cbv$ order, therefore (by assumption) it terminates. Because there are no $\tY$-reductions in the evaluation of $F$, it can't reach any other term of the form $F\ \underline{n_k'}\ X$, therefore $k$ never changes until this whole subexpression reaches a number, at which point $k$ decreases by 1 and the next $F$ is evaluated.

### 2021-01-18 LICS21/antitone.tex
Minor fixes

Define an \emph{antitone sparse ranking function} on a closed term $M$ to be a partial function $F : Rch(M) \rightharpoonup \mathbb{R}$ such that for some antitone function $\epsilon : \mathbb{R}_{\geq 0} \to \mathbb{R}_{>0}$:
\begin{inparaenum}[(i)]
\item $f(N) \geq 0$ for all $N \in \dom(f)$,
\item $M \in \dom(f)$,
\item for any $N \in \dom(f)$, evaluation of $N$ will eventually reach some $O$ which is either a value or in $\dom(f)$, and $f(N) \geq \mathbb E [f(O) + \epsilon(f(O)) \times \text{the number of $\tY$-reduction steps from $N$ to $O$}]$ (where $f(O)$ is taken to be 0 if $O$ is a value outside of $\dom(f)$).
\end{inparaenum}
\iffalse
\begin{itemize}
\item $f(N) \geq 0$ for all $N$ where $f$ is defined.
\item $f$ is defined at $M$.
\item For any $N$ in the domain of definition of $f$, evaluation of $N$ will eventually reach some $O$ which is either a value or in the domain of definition of $f$, and $f(N) \geq \mathbb E [f(O) + \epsilon(f(O)) \times \text{the number of $\tY$-reduction steps from $N$ to $O$}]$ (where $f(O)$ is taken to be 0 if $O$ is a value outside of the domain of $f$).
\end{itemize}
\fi
\begin{restatable}[Sparse function]{theorem}{AntitonePartial}  \label{thm:antitone partial implies rankable}
  Every antitone sparse ranking function is a restriction of an antitone ranking function.
\end{restatable}

@@

\begin{example}[Continuous random walk]\label{ex:raven complex}
In PPCF we can construct a function whose argument changes by a random amount at each recursive call: $\Theta \, 10$ where
\[
%\underbrace{\big
\Theta := \tY \, \lambda f x . \tif{x \leq 0}{0}{f(x - \tsample)}
%}_{\Theta} \, 10
\]

For a more complex (not $\tY$-PAST) example, consider the following ``continuous random walk'': $\Xi \, 10$ where
\[
%\underbrace{\big(
\Xi := \tY \, f \, x . \tif{x \leq 0}{0}{f(x - \tsample + 1/2)} 
%\big)}_{\Xi} \, 10
\]
\end{example}

\begin{example}[Fair-in-the-limit random walk]
\label{ex:Fair-in-the-limit random walk}\cite[\S 5.3]{DBLP:journals/pacmpl/McIverMKK18}
\[
\big
(\tY \, f \, x . 
\tif{x \leq 0}{0}{f(x - 1) \oplus_{\frac{x}{2x+1}} f(x + 1)} \big)
\, 10
\]
\end{example}

\begin{example}[Escaping spline]
\label{ex:escaping spline}\cite[\S 5.4]{DBLP:journals/pacmpl/McIverMKK18}
\[
%\underbrace{
\big(\tY \, f \, x . \,
0 \oplus_{\frac{1}{x+1}} f(x + 1) 
\big)
%}_{\Xi} 
\, 10
\]
\end{example}

>>>>

\begin{definition}
Define an \emph{antitone sparse ranking function} on a closed term $M$ to be a partial function $F : Rch(M) \rightharpoonup \mathbb{R}$ such that for some antitone function $\epsilon : \mathbb{R}_{\geq 0} \to \mathbb{R}_{>0}$:
\begin{inparaenum}[(i)]
\item $f(N) \geq 0$ for all $N \in \dom(f)$,
\item $M \in \dom(f)$,
\item for any $N \in \dom(f)$, evaluation of $N$ will eventually reach some $O$ which is either a value or in $\dom(f)$, and $f(N) \geq \mathbb E [f(O) + \epsilon(f(O)) \times \text{the number of $\tY$-reduction steps from $N$ to $O$}]$ (where $f(O)$ is taken to be 0 if $O$ is a value outside of $\dom(f)$).
\end{inparaenum}
\end{definition}

\begin{restatable}[Sparse function]{theorem}{AntitonePartial}  \label{thm:antitone partial implies rankable}
  Every antitone sparse ranking function is a restriction of an antitone ranking function.
\end{restatable}

@@

\begin{example}[Continuous random walk]\label{ex:raven complex}
In PPCF we can construct a function whose argument changes by a random amount at each recursive call: $\Theta \, \underline{10}$ where
\[
%\underbrace{\big
\Theta := \tY \, \lambda f x .\ \tif{x \leq 0}{\underline 0}{f(x - \tsample)}
%}_{\Theta} \, 10
\]

For a more complex (not $\tY$-PAST) example, consider the following continuous random walk: $\Xi \, \underline{10}$ where
\[
\Xi := \tY \, \lambda f x .\ \tif{x \leq 0}{\underline 0}{f(x - \tsample + 1/2)} 
\]
\end{example}

\begin{example}[Fair-in-the-limit random walk]
\label{ex:Fair-in-the-limit random walk}\cite[\S 5.3]{DBLP:journals/pacmpl/McIverMKK18}
\[
\big
(\tY\ \lambda f x .\ 
\tif{x \leq 0}{\underline 0}{f(x - 1) \oplus_{\frac{x}{2x+1}} f(x + 1)} \big)
\, \underline{10}
\]
\end{example}

\begin{example}[Escaping spline]
\label{ex:escaping spline}\cite[\S 5.4]{DBLP:journals/pacmpl/McIverMKK18}
\[
%\underbrace{
\big(\tY\ \lambda f x . \,
\underline 0 \oplus_{\frac{1}{x+1}} f(x + 1) 
\big)
%}_{\Xi} 
\, \underline{10}
\]
\end{example}

### 2021-01-18 LICS21/applications.tex
Introduction:ongoing Reformatted last antitone sparse ranking function in appendix
Minor fixes

It is possible to reduce this problem by labelling the samples in the trace by something other than the order in which they are used. If the labels of samples in the trace correspond closely to the roles that the samples play in the program execution, this will increase the efficiency of the inference algorithm. 
In \cite{wingate2011lightweight}, for example, something rather like a stack trace is used, although this assdressing scheme turns out to have various problems \cite{kiselyov2016problems,Hur2015a}. 
Potential positions could also be used for a similar purpose, although in practice, a more compact representation than an entire skeletal reduction sequence would be necessary. 
%\lo{It would be very interesting (and rewarding, I believe) to work out the details. There is a formalisation of LMH for Idealised Anglican in \Cref{fnote:lmh}. SPCF is a much nicer probabilistic programming language (in purified / idealised form).}
Selecting an appropriate addressing scheme can also be important in the context of variational inference \cite[\S 6.2.2]{paige2016automatic}.

>>>>

It is possible to reduce this problem by labelling the samples in the trace by something other than the order in which they are used. If the labels of samples in the trace correspond closely to the roles that the samples play in the program execution, this will increase the efficiency of the inference algorithm. 
In \cite{wingate2011lightweight}, for example, something rather like a stack trace is used, although this addressing scheme turns out to have various problems \cite{kiselyov2016problems,Hur2015a}. 
Potential positions could also be used for a similar purpose, although in practice, a more compact representation than an entire skeletal reduction sequence would be necessary. 
%\lo{It would be very interesting (and rewarding, I believe) to work out the details. There is a formalisation of LMH for Idealised Anglican in \Cref{fnote:lmh}. SPCF is a much nicer probabilistic programming language (in purified / idealised form).}
Selecting an appropriate addressing scheme is also important for variational inference \cite{paige2016automatic}.
%\cite[\S 6.2.2]{paige2016automatic}.

### 2021-01-18 LICS21/confluent-proofs.tex
Introduction:ongoing Reformatted last antitone sparse ranking function in appendix
minor
Minor fixes

@@

@@

Providing a ranking function of some sort for this term in a similar level of detail would have been rather more complicated using only the standard reduction strategy. The recursion in $\Theta$ would have had to be evaluated separately for every time it was used, so there would have been more (and more complex) terms to consider in the reduction sequence. Also, the number of $\tY$-reductions from $P[\underline n]$ would have been $n^2 + 2n + 1$ instead of $2n + 2$, therefore the expected number of $\tY$ reductions starting from $M$ would be infinite, i.e.~$M$ is not $\tY$-PAST, therefore it is not even rankable. It would be possible to define an antitone sparse ranking function for it, but it is much more difficult to construct:
\begin{align*}
& M \mapsto 1.47 \\
& P[\underline n] \mapsto 1 + \ln((n+1)n + 1) \\
& \Xi[\lambda x. \Theta[x] \underline n] \underline n \mapsto 1 + \ln((n+1)n + 1) \\
& \underline{n^n} + \dots + \underline{k_1^n} + \\
& \Xi[\lambda x. \Theta[x] \underline n] (\underline{k_1} - 1) \mapsto 1 + \ln((n+1)(k_1-1)+1) \\
& \underline{n^n} + \dots + \underline{(k_1+1)^n} + \Theta[\underline{k_1}] \underline{n} + \\
& \Xi[\lambda x. \Theta[x] \underline n] (\underline{k_1} - 1) \mapsto 1 + \ln((n+1)k_1 + 1) \\
& \underline{n^n} + \dots + \underline{(k_1+1)^n} + \\
& \underbrace{\underline{k_1} \times \dots \times \underline{k_1} \times}_{n-k_2+1} \Theta[\underline{k_1}] (\underline{k_2} - 1) + \\
& \Xi[\lambda x. \Theta[x] \underline n] (\underline{k_1} - 1) \mapsto 1 + \ln((n+1)k_1 + k_2 + 1) \\
& \underline{\sum_{k=0}^n k^n} \mapsto 0.
\end{align*}

>>>>

It is also true that the distributions of terms produced by the standard semantics and by the reduction strategy $\cbv$ are the same, for much the same reason, although that is not relevant to our main results.

@@

It is not quite true, however, that all reduction strategies produce the same values (barring some reduction strategies terminating where others don't), because it is possible for a reduction strategy to overshoot the value that the standard semantics produces by, for example, performing reductions inside of a lambda. Even for the same value of the samples, the final values may be different. A more restricted result is true, however, that if a term terminates with some reduction strategy $f$ and samples $s$, the value that $\Rightarrow_{\cbv}$ terminates with can reduce via $\Rightarrow$ to the value that $\Rightarrow_f$ terminates with (as a corollary to this proof).

@@

Providing a ranking function of some sort for this term in a similar level of detail would have been rather more complicated using only the standard reduction strategy. The recursion in $\Theta$ would have had to be evaluated separately for every time it was used, so there would have been more (and more complex) terms to consider in the reduction sequence. Also, the number of $\tY$-reductions from $P[\underline n]$ would have been $n^2 + 2n + 1$ instead of $2n + 2$, therefore the expected number of $\tY$ reductions starting from $M$ would be infinite, i.e.~$M$ is not $\tY$-PAST, therefore it is not even rankable. It would be possible to define an antitone sparse ranking function for it, but it is much more difficult to construct:
\iffalse
\begin{align*}
& M \mapsto 1.47 \\
& P[\underline n] \mapsto 1 + \ln((n+1)n + 1) \\
& \Xi[\lambda x. \Theta[x] \underline n] \underline n \mapsto 1 + \ln((n+1)n + 1) \\
& \underline{n^n} + \dots + \underline{k_1^n} + \\
& \Xi[\lambda x. \Theta[x] \underline n] (\underline{k_1} - 1) \mapsto 1 + \ln((n+1)(k_1-1)+1) \\
& \underline{n^n} + \dots + \underline{(k_1+1)^n} + \Theta[\underline{k_1}] \underline{n} + \\
& \Xi[\lambda x. \Theta[x] \underline n] (\underline{k_1} - 1) \mapsto 1 + \ln((n+1)k_1 + 1) \\
& \underline{n^n} + \dots + \underline{(k_1+1)^n} + \\
& \underbrace{\underline{k_1} \times \dots \times \underline{k_1} \times}_{n-k_2+1} \Theta[\underline{k_1}] (\underline{k_2} - 1) + \\
& \Xi[\lambda x. \Theta[x] \underline n] (\underline{k_1} - 1) \mapsto 1 + \ln((n+1)k_1 + k_2 + 1) \\
& \underline{\sum_{k=0}^n k^n} \mapsto 0.
\end{align*}
\fi
\begin{align*}
M & \mapsto 1.47 \\
P[\underline n] & \mapsto 1 + \ln((n+1)n + 1) \\
\Xi[\lambda x. \Theta[x] \, \underline n] \, \underline n & \mapsto 1 + \ln((n+1)n + 1) \\
%\underline{n^n} + \dots + \underline{k_1^n} + \\
%\Xi[\lambda x. \Theta[x] \underline n] (\underline{k_1} - 1) & \mapsto 1 + \ln((n+1)(k_1-1)+1) \\
\left.\begin{array}{r}
\underline{n^n} + \dots + \underline{k_1^n} + {}\\
\Xi[\lambda x. \Theta[x] \underline n] (\underline{k_1} - 1) 
\end{array}\right\} 
& \mapsto 1 + \ln((n+1)(k_1-1)+1) \\
%\underline{n^n} + \dots + \underline{(k_1+1)^n} + \Theta[\underline{k_1}] \underline{n} + \\
%\Xi[\lambda x. \Theta[x] \underline n] (\underline{k_1} - 1) & \mapsto 1 + \ln((n+1)k_1 + 1) \\
\left.\begin{array}{r}
\underline{n^n} + \dots + \underline{(k_1+1)^n} +  {}\\
\Theta[\underline{k_1}] \, \underline{n} + \Xi[\lambda x. \Theta[x] \, \underline n] (\underline{k_1} - 1) 
\end{array}\right\}
& \mapsto 1 + \ln((n+1)k_1 + 1) \\
%\underline{n^n} + \dots + \underline{(k_1+1)^n} + \\
%\underbrace{\underline{k_1} \times \dots \times \underline{k_1} \times}_{n-k_2+1} \Theta[\underline{k_1}] (\underline{k_2} - 1) + \\
%\Xi[\lambda x. \Theta[x] \underline n] (\underline{k_1} - 1) & \mapsto 1 + \ln((n+1)k_1 + k_2 + 1) \\
\left.\begin{array}{r}\underline{n^n} + \dots + \underline{(k_1+1)^n} + {} \\
\underbrace{\underline{k_1} \times \dots \times \underline{k_1} \times}_{n-k_2+1} \Theta[\underline{k_1}] (\underline{k_2} - 1) + {} \\
\Xi[\lambda x. \Theta[x] \, \underline n] (\underline{k_1} - 1)
\end{array}\right\} & \mapsto 1 + \ln((n+1)k_1 + k_2 + 1) \\
\underline{\textstyle\sum_{k=0}^n k^n} & \mapsto 0.
\end{align*}

### 2021-01-18 LICS21/confluent.tex
Minor fixes

\begin{example}[Labelling samples by terms]
Consider the terms
\begin{align*}
A[x] &= \tif{\tif{x>0}{I}{I} \, \tsample - 0.5 > 0}{0}{\Omega}\\
B &= \tif{\tsample - 0.5 > 0}{0}{\Omega}
\end{align*}
If terms rather than skeletons were used to label samples, the set of modified traces where $A[\tsample]$ terminates would be
\[
\begin{array}{l}
\bigcup_{r \in [0,1]} \big\{s \mid\ s(A[\tsample], \mathsf{if}_1;\underline{-}_1;@_1;\mathsf{if}_1) = r, \\
\qquad\qquad\qquad s(A[\tsample] \to A[\underline r] \to^* B, \mathsf{if}_1 ; \underline{-}_1) > 0.5 \big\}.
\end{array}
\]
\iffalse
\lo{The occurrence of $L_s(M)$ in the set comprehension below should be $L_0(M)$?}

@@

        \item $N \mid \alpha = \tY (\lambda x. A)$, $A$ reduced at $\beta$ is $A'$, $(\gamma_i)_i$ is a list of all the positions where $A' \mid \gamma  = x$, ordered from left to right, $O_1$ is $N$ reduced first at $\alpha;\tY;\lambda;\beta$ then at $\alpha$, and $O_2$ is $N$ reduced first at $\alpha$ then at $\alpha;\lambda;@_1;\gamma_i;Y;\lambda;\beta$ for each $i$ in order where $\gamma_i$ is left of $\beta$ then at $\alpha;\lambda;@_1;\beta$ then at $\alpha;\lambda;@_1;\gamma_i;Y;\lambda;\beta$ for the remaining values of $i$.
    \end{enumerate}
    (in which case $O_1$ and $O_2$ are equal as skeletons, but with different reduction sequences), $O_1'$ and $O_2'$ are the results of applying some reduction sequence to each of $O_1$ and $O_2$ (the same reductions in each case, which is always possible because they're equal terms), and $\delta$ is a position in $O_1'$ (or equivalently $O_2'$), then $(O_1',\delta) \sim_c (O_2',\delta)$.
\end{asparaenum}
The $\sim_c$-rules are illustrated in \Cref{sim-diagram}.
\end{definition}

@@

It is also true that the distributions of terms produced by the standard semantics and by the reduction strategy $\cbv$ are the same, for much the same reason, although that is not relevant to our main results.

@@

It is not quite true, however, that all reduction strategies produce the same values (barring some reduction strategies terminating where others don't), because it is possible for a reduction strategy to overshoot the value that the standard semantics produces by, for example, performing reductions inside of a lambda. Even for the same value of the samples, the final values may be different. A more restricted result is true, however, that if a term terminates with some reduction strategy $f$ and samples $s$, the value that $\Rightarrow_{\cbv}$ terminates with can reduce via $\Rightarrow$ to the value that $\Rightarrow_f$ terminates with (as a corollary to the proof of \Cref{thm:CbvIsTerminatingest}).

>>>>

\begin{example}[Labelling samples by terms]
Consider the terms
\begin{align*}
A[x] &= \tif{\tif{x>0}{I}{I} (\lambda y.\tsample)\ \underline 0 - \underline{0.5} > 0}{\underline 0}{\Omega}\\
B &= \tif{\tsample - \underline{0.5} > 0}{\underline 0}{\Omega}
\end{align*}
If terms rather than skeletons were used to label samples, the set of modified traces where $A[\tsample]$ terminates would be
\[
\begin{array}{l}
\bigcup_{r \in [0,1]} \big\{s \mid\ s(A[\tsample], \mathsf{if}_1;\underline{-}_1;@_1;@_1;\mathsf{if}_1) = r, \\
\qquad\qquad\qquad s(A[\tsample] \to A[\underline r] \to^* B, \mathsf{if}_1 ; \underline{-}_1) > 0.5 \big\}.
\end{array}
\]
\iffalse
\lo{The occurrence of $L_s(M)$ in the set comprehension below should be $L_0(M)$?}

@@

        \item $N \mid \alpha = \tY (\lambda x. A)$, $A$ reduced at $\beta$ is $A'$, $(\gamma_i)_i$ is a list of all the positions where $A' \mid \gamma  = x$, ordered from left to right, $O_1$ is $N$ reduced first at $\alpha;\tY;\lambda;\beta$ then at $\alpha$, and $O_2$ is $N$ reduced first at $\alpha$ then at $\alpha;\lambda;@_1;\gamma_i;Y;\lambda;\beta$ for each $i$ in order where $\gamma_i$ is left of $\beta$ then at $\alpha;\lambda;@_1;\beta$ then at $\alpha;\lambda;@_1;\gamma_i;Y;\lambda;\beta$ for the remaining values of $i$.
    \end{enumerate}
    (in which case $O_1$ and $O_2$ are equal as skeletons, but with different reduction sequences), $O_1'$ and $O_2'$ are the results of applying some reduction sequence to each of $O_1$ and $O_2$ (the same reductions in each case, which is always possible because they're equal skeletons), and $\delta$ is a position in $O_1'$ (or equivalently $O_2'$), then $(O_1',\delta) \sim_c (O_2',\delta)$.
\end{asparaenum}
The $\sim_c$-rules are illustrated in \Cref{sim-diagram}.
\end{definition}

@@

@@

### 2021-01-18 LICS21/intro.tex
Minor fixes

In this paper we study a central property of probabilistic programs: \emph{termination}. 
A program is \emph{almost surely terminating} (AST) if a run of it terminates with probability one.
When a probabilistic program implements a solution to an algorithmic problem, we naturally require the computation to terminate with a high probability, usually $1$.
In probabilistic programming, guarantees of AST are equally important. 
Indeed, it is standard for designers and implementors of probabilistic programming systems to regard non-AST programs as defining invalid models, and hence inadmissible \cite{Rainforth2017,DBLP:conf/uai/GoodmanMRBT08}.
(Yet none of these systems provides any support for the development or verification of AST programs.)

>>>>

In this paper we study a central property of probabilistic programs: \emph{termination}.
When a probabilistic program implements a solution to an algorithmic problem, we naturally require the computation to terminate with probability $1$, in which case the program is called \emph{almost surely terminating} (AST).
Indeed, it is standard for designers and implementors of probabilistic programming systems to regard non-AST programs as defining invalid models, and hence inadmissible \cite{Rainforth2017,DBLP:conf/uai/GoodmanMRBT08}.
(Yet none of these systems provides any support for the development or verification of AST programs.)
